Verification of concurrent design patterns with data

### 2019-02-22 2018-Arch-pNets/arch-pNets.tex
Nettoye presque toutes le notes...

In this paper, we  use  a refined
version of the Failure Monitor architecture
from~\cite{CubETH-case-study} as a running example.  Although \fig{schema:ArchFailure:BIP}
shows the full definition of this architecture, we will explain its
various elements progressively.
\fig{schema:ArchFailure:BIP} shows  components
%% The
%% component $\NameOprnd$ is an \emph{operand} of the architecture, \ie
%% it only has its interface defined: $(\emptyset, \{\PortFinish,
%% \PortResume, \PortFail\}, \export[\emptyset])$, with
%% $\export[\emptyset](p) = \emptyset$, for all $p \in \{\PortFinish,
%% \PortResume, \PortFail\}$.  
%% The other two components,
%
$\NameTimer$(imer) and $\NameCtrl$(ontrol), with interfaces
$\btuple{
  \{t, z\},
  \{\PortTick, \PortCancel, \PortTOT\},
  \bigl\{\!\PortTick \mapsto \{t, z\}\!\bigr\}
}$,
and
$\btuple{
  \{\mathit{zone}\},
  \{\PortReset, \PortFail, \PortResume, \PortTOC\},
  \bigl\{\!\PortFail \mapsto \{\mathit{zone}\}\!\bigr\}
}$
\noteEM{Attention, bug dans la figure 2, ou tu mets $reset\mapsto \mathit{zone}$}\noteSB{Done!}respectively.
%
%% \[
%% \begin{array}{@{}c@{\hspace{3mm}}c@{}}
%%   \bigl(
%%   \{t, z\},
%%   \{\PortTick, \PortCancel, \PortTOT\},
%%   \export[\NameTimer]
%%   \bigr)\,,
%%   %% \qquad
%%   &
%%   \bigl(
%%   \{\mathit{zone}\},
%%   \{\PortReset, \PortFail, \PortResume, \PortTOC\},
%%   \export[\NameCtrl]
%%   \bigr)\,,
%%   \\
%%   \export[\NameTimer](p) =
%%   \begin{cases}
%%     \{t, z\}, & \text{if } p = \PortTick,\\
%%     \emptyset, & \text{otherwise,}
%%   \end{cases}
%%   %% \]
%%   %% and
%%   %% \[
%%   %% \quad
%%   &
%%   \export[\NameCtrl](p) =
%%   \begin{cases}
%%     \{\mathit{zone}\}, & \text{if } p = \PortFail,\\
%%     \emptyset, & \text{otherwise.}
%%   \end{cases}
%% \end{array}
%% \]
%
%% , where $\export[\NameCtrl](\PortFail) =
%% \{\mathit{zone}\}$ and $\export[\NameCtrl](p) = \emptyset$, for all
%% other ports $p$; $\export[\NameTimer](\PortTick) = \{t, z\}$ and
%% $\export[\NameTimer](p) = \emptyset$, for all other ports $p$.
Variable $t$ is implicitly assumed to be an integer.  Variables $z$
and $\mathit{zone}$ are  integer intervals of
the form $[l,u]$.  Both types are implicitly equipped with a
natural ordering: increasing integers and interval inclusion.

@@

\addSB{Intuitively, an interaction can be fired if all the involved
  components are ready to fire their corresponding transitions.  The
  other components do not change their states.  Both the interaction
  guard and those of the participating transitions must be satisfied.
  The update assignment of the interaction is executed first, followed
  by those of the components.}

@@

\addSB{In addition to interaction models, BIP relies on 
  \emph{priority models} that impose  a strict partial order on
  interactions.  Formal definitions are provided in \app{arch:appendix}.  Intuitively, an interaction can be fired only if all
  the higher-priority interactions available in the current state are
  disabled by their respective guards.
%
In the next sections, we will implicitly assume application of the  \emph{maximal
  progress} priority $\mu$, where $(a, g, u) \prec_\mu (b, h, w)$ iff  $a \subset b$ and $a \neq b$.  For instance, the
port $\NameTimer.\PortTick$ will never fire alone if the port $\PortFail$ is also enabled.
In \secn{conclusion}, we discuss the generalisation of the
obtained results to other priority models.
}
%% Below, we will assume that the maximal progress priority $\mu$ is
%% applied after each composition with an interaction model.

@@

\label{property:reset}
Although we define properties as state predicates, any
appropriate logic can be used to specify them.  For instance, the
property \addSB{\emph{``There is always a possibility to reset the system
  after a single failure''} (\ie without
  additional failures having to occur in the meantime)} enforced by the
Failure Monitor architecture can be specified using CTL as \noteSB{Removed $\lnot \PortResume$, because not needed \emph{and} could not reformulate nicely in English.}\addSB{$\AG
\bigl(\PortFail \rightarrow \EX \EU[]{\lnot \PortFail}{
  \PortReset}\bigr)$}.  Additional properties are provided in
\app{properties}.
%% \todoSB{Cite the original paper somewhere.}
An architecture enforces
its characteristic property on  its operand components.
From this point of view, the set of coordinating components is
not relevant, neither are their states.  Thus, 
properties enforced by architectures only involve
the unrestricted composition of the operands:

@@

\addSB{%
%
\Thm{combining} implies that safe BIP systems can be designed
\emph{compositionally}.  Indeed, it is sufficient to verify that
1)~the applied architectures do enforce their characteristic
properties and 2)~that they are pairwise upwards compatible.  To
a large extent, item~2, \ie verification of the upwards
compatibility, can be carried out syntactically by analysing the
structure of the connectors defining the interaction models:
maximal progress is mostly relevant for connectors involving
triggers; whenever a trigger port is not common to both
architectures, upwards \emph{in}compatibility cannot arise.  The
next section is devoted to the encoding of architectures into
pNets, which addresses item~1 by symbolic verification.
%
}

@@

\paragraph{Encoding the interaction model}
%% First of all, recall that $\nopri[A, \partition ] \bydef{=}
%% \pNetTuple{(\nopri[C])^{C \in \cC}, \partition, \nopri[\Gamma]}$.
%% Thus, 
The holes in $\nopri[A, \partition]$ are indexed by the elements of
the partition \partition.  For the encoding of our running example, we
take $\partition = \bigl\{\{\PortFail, \PortResume\},\{\PortAsk,
\PortReset\}\bigr\}$.  This corresponds to the intuition that the
dangling ports {\PortFail} and {\PortResume} will be provided by a \noteLH{monitored ou monitoring???}
monitored component, whereas {\PortAsk} and {\PortReset} correspond to
the actions provided by the ``environment'' (other components of the
system) that are invoked in case of a persistent failure.  As for
the encoding of the coordinators, in the synchronisation vectors of
\nopri[\Gamma], we will associate Boolean values to the actions
corresponding to these ports.

@@

\addSB{We encode all interactions of a connector in one synchronisation vector.  This will allow us to also encode the maximal progress priority model.}
We use
the additional Boolean values associated to each port by the encoding
of coordinator components.  For example, observe that the three ports in 
the connector
{\NameTimer.\PortTick
  \trigsynch (\PortFail \trigsynch \NameCtrl.\PortFail)}
form a ``causality chain'': {\NameCtrl.\PortFail} can only
participate in an interaction if the dangling port {\PortFail}
participates, which in turn can only happen if
{\NameTimer.\PortTick} does so.  These dependencies can be rewritten
as Boolean implications $\NameCtrl.\PortFail \Rightarrow \PortFail$
and $\PortFail \Rightarrow \NameTimer.\PortTick$.
%
The conjunction of these two implications can be used as a guard for
the synchronisation vector encoding this connector.

@@

\noteEM{Dans le paragraph precedent, tu resume le cas des connecteurs sans data, dans les prochains, tu formalise avec data, mais pour le lecteur ``casual'', le changement n'est pas evident... }
Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, \addSB{the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
  and the encoding of $\Gamma$ is} $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  {\color{blue}\alpha \bigl(b_p^{p \in P_x}\bigr)}
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%
with $b_p^{p \in P_x}$ fresh Boolean variables, $\alpha$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of all $p$ by corresponding variables $b_p$.

\addSB{%
%
For the sake of clarity, we simplify the case study encoding in
\fig{FailureTimer:pNet} by reusing the port names of the original
architecture instead of fresh names $\alpha$.  This is made
possible by the fact that each synchronisation vector involves at
most one action of interest (see the properties in
\app{properties}).
%% In the general encoding, the global actions of the
%% synchronisation vectors should be fresh action names.
%% However,
%% for the properties considered in our case study, we only need to
%% exhibit the actions of the holes.  Furthermore, each
%% synchronisation vector involves at most one such action.
%% This
%% allows us to reuse them as the global actions of the
%% synchronisation vectors, simplifying the translation of
%% properties to be verified (\cf \app{properties}).
%
}\noteLH{ce
  paragraphe sonne vraiment note de rapport de recherche, tu
  arrives a lui donner un peu d'abstraction/de recul?}\noteSB{Est-ce que c'est mieux comme \c{c}a ? OUI -- Ludo}

@@

\paragraph{Practical experiments}\noteSB{J'ai repris ce paragraphe}
\noteLH{j ai nomme ce paragraphe}
The above encoding provides a mechanism for the symbolic verification
of architectures using our existing tool
\cite{HMZ-FORTE2016QBMZ-AVOCS18} to compute the open automaton
semantics of an open pNet. This tool computes open transitions from
pLTS behaviours and synchronisation vectors of the pNet, then uses an
SMT engine to check satisfiability of their predicates, minimizing the
size of the resulting automaton.
%% The encoding into the SMT language involves the definition of the
%% action and data domains and operators, and the axiomatisation of
%% domain-dependant properties, to allow the SMT engine to be as precise
%% as possible (even if it cannot be complete).
%% The resulting open automaton will later be subject to (symbolic)
%% model-checking and equivalence checking algorithms, themselves
%% requiring automatic proof assistance. \todoEM{trop flou pour le
%%   moment, mais je n'ose pas dire des trucs trop precis... meme si on a
%%   deja une implem de la FH-bisim, elle n'est pas validee!}

In \fig{ArchFailure:OA} we show the full open automaton
obtained from the pNet in \fig{FailureTimer:pNet}.  Due to space limitations, we do not show the
details of the open transitions, but only the assignments of state
variables, some useful parts of the predicates, and the resulting
action; full details can be found in 
\app{full-results}.
%
\noteSB{Removed ``As an example...'', since that \ex{TimerOT3} was
  simplified and does not correspond to the real OT.}
%% As an example, $OT_3$ is the one
%% explained in \ex{TimerOT3} (page~\pageref{ex:TimerOT3}).
This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OT 2 and 3 for
{\PortFail}). \add{It looks like} the rough behaviour is
reasonable, although a detailed study shows that the property stated
in page~\pageref{property:reset} does not hold, because of the
{\PortFail} loops on states $11$ and $12$. This is because we did not
use the maximal progress assumption here. If we do, we get the
corrected behaviour in \fig{ArchFailure:OA-MaxProgress}
(\app{full-results}), in which $OT_7$ has disappeared, and $OT_2$ and
$OT_{10}$ have been restricted to $b0=\false$. This one verifies all
the properties listed in \app{properties}.

@@

\noindent
\add{The following appendices are included for convenience and would be integrated into a research report if the paper is accepted for publication. The first appendix provides previously published complementary information, whereas the subsequent ones give additional formal details complementing architecture definitions, proofs of theorems, and the case study.}

@@

\noindent
\verbatimfont{\small}
\begin{verbatim}
SV0: <tick(z, t, z1), fail(zone, b1), fail(b0), _> --> fail(b0),
         [(b1->b0)/\(z1=z |Intersection| ((b1?zone:TopInt)+t))]
SV1: <cancel(b1), resume(b2), resume, _> --> resume, (b1=b2)
SV2: <timeout, timeout, _, ask> --> ask
SV3: <_, reset, _, reset> --> reset
SV4: <init, init, _, _> --> init
pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [(t<GET_MAX z)], { t := (t+1) z := z1}
- Transition: t1---cancel(false)-->t1, {}
- Transition: t1---cancel(true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := (t+1)}
pLTS: Control
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false)-->s1, {}
- Transition: s1---fail(zone, false)-->s1, {}
- Transition: s1---fail(zone, true)-->s2, {}
- Transition: s2---fail(zone, false)-->s2, {}
- Transition: s2---resume(false)-->s2, {}
- Transition: s2---resume(true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false)-->s3, {}
- Transition: s3---resume(false)-->s3, {}
- Transition: s3---reset-->s1, {}
Holes: B
Holes: E
\end{verbatim}

@@

\verbatimfont{\small}
\begin{verbatim}
   {t0---init-->t1, s0---init-->s1},   {},  [true],
   {t := 0, z := TopInt, zone := minZone,maxZone}
OT1 ------------------------------------------------------------
    <t0_s0>-----init----><t1_s1>

@@

INFO  BIPArchFailureTimerMax - Total: 12 SATISFIABLE OTs
INFO  BIPArchFailureTimerMax -        235 UNSATISFIABLE OTs
INFO  BIPArchFailureTimerMax - Total Time: 1293 ms
\end{verbatim}

@@

\medskip
\noindent
$  OT_1  = \openrule
{\emptyset, \{t:=0, z:=TopInt, zone:=[minZone,maxZone]\} }
{\ostate{00} \OTarrow{init} \ostate{11}}
$

@@

  \subsubsection{Properties}

@@

\small\begin{verbatim}
SV0: <tick(z, t, z1), fail(zone, bf, b1), fail(b0), _> --> fail(b0),
         [(b1=>b0)/\((bf/\b0)=>b1)/\(z1=(z |Intersection| ((b1?zone:TopInt)+t)))]
SV1: <cancel(bc, b1), resume(br, b2), resume, _>-->resume, (b1=b2)/\((bc/\br)=>(b1/\b2))
SV2: <timeout, timeout, _, ask> --> ask
SV3: <_, reset, _, reset> --> reset
SV4: <init, init, _, _> --> init

pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
- Transition: t1---cancel(true, false)-->t1, {}
- Transition: t1---cancel(true, true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}
pLTS: Control
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false, false)-->s1, {}
- Transition: s1---fail(zone, true, false)-->s1, {}
- Transition: s1---fail(zone, true, true)-->s2, {}
- Transition: s2---fail(zone, false, false)-->s2, {}
- Transition: s2---resume(true, false)-->s2, {}
- Transition: s2---resume(true, true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false, false)-->s3, {}
- Transition: s3---resume(false, false)-->s3, {}
- Transition: s3---reset-->s1, {}
\end{verbatim}

@@

\subsubsection{Scaling-up: composing architectures}

>>>>

In this paper, we  use  a refined
version of the Failure Monitor architecture
from~\cite{CubETH-case-study} as a running example.  Although \fig{schema:ArchFailure:BIP}
shows the full definition of this architecture, we will explain its
various elements progressively.
\fig{schema:ArchFailure:BIP} shows  components
%% The
%% component $\NameOprnd$ is an \emph{operand} of the architecture, \ie
%% it only has its interface defined: $(\emptyset, \{\PortFinish,
%% \PortResume, \PortFail\}, \export[\emptyset])$, with
%% $\export[\emptyset](p) = \emptyset$, for all $p \in \{\PortFinish,
%% \PortResume, \PortFail\}$.  
%% The other two components,
%
$\NameTimer$(imer) and $\NameCtrl$(ontrol), with interfaces
$\btuple{
  \{t, z\},
  \{\PortTick, \PortCancel, \PortTOT\},
  \bigl\{\!\PortTick \mapsto \{t, z\}\!\bigr\}
}$,
and
$\btuple{
  \{\mathit{zone}\},
  \{\PortReset, \PortFail, \PortResume, \PortTOC\},
  \bigl\{\!\PortFail \mapsto \{\mathit{zone}\}\!\bigr\}
}$
%\noteEM{Attention, bug dans la figure 2, ou tu mets $reset\mapsto \mathit{zone}$}\noteSB{Done!}
respectively.
%
%% \[
%% \begin{array}{@{}c@{\hspace{3mm}}c@{}}
%%   \bigl(
%%   \{t, z\},
%%   \{\PortTick, \PortCancel, \PortTOT\},
%%   \export[\NameTimer]
%%   \bigr)\,,
%%   %% \qquad
%%   &
%%   \bigl(
%%   \{\mathit{zone}\},
%%   \{\PortReset, \PortFail, \PortResume, \PortTOC\},
%%   \export[\NameCtrl]
%%   \bigr)\,,
%%   \\
%%   \export[\NameTimer](p) =
%%   \begin{cases}
%%     \{t, z\}, & \text{if } p = \PortTick,\\
%%     \emptyset, & \text{otherwise,}
%%   \end{cases}
%%   %% \]
%%   %% and
%%   %% \[
%%   %% \quad
%%   &
%%   \export[\NameCtrl](p) =
%%   \begin{cases}
%%     \{\mathit{zone}\}, & \text{if } p = \PortFail,\\
%%     \emptyset, & \text{otherwise.}
%%   \end{cases}
%% \end{array}
%% \]
%
%% , where $\export[\NameCtrl](\PortFail) =
%% \{\mathit{zone}\}$ and $\export[\NameCtrl](p) = \emptyset$, for all
%% other ports $p$; $\export[\NameTimer](\PortTick) = \{t, z\}$ and
%% $\export[\NameTimer](p) = \emptyset$, for all other ports $p$.
Variable $t$ is implicitly assumed to be an integer.  Variables $z$
and $\mathit{zone}$ are  integer intervals of
the form $[l,u]$.  Both types are implicitly equipped with a
natural ordering: increasing integers and interval inclusion.

@@

  Intuitively, an interaction can be fired if all the involved
  components are ready to fire their corresponding transitions.  The
  other components do not change their states.  Both the interaction
  guard and those of the participating transitions must be satisfied.
  The update assignment of the interaction is executed first, followed
  by those of the components.

@@

In addition to interaction models, BIP relies on 
  \emph{priority models} that impose  a strict partial order on
  interactions.  Formal definitions are provided in \app{arch:appendix}.  Intuitively, an interaction can be fired only if all
  the higher-priority interactions available in the current state are
  disabled by their respective guards.
%
In the next sections, we will implicitly assume application of the  \emph{maximal
  progress} priority $\mu$, where $(a, g, u) \prec_\mu (b, h, w)$ iff  $a \subset b$ and $a \neq b$.  For instance, the
port $\NameTimer.\PortTick$ will never fire alone if the port $\PortFail$ is also enabled.
In \secn{conclusion}, we discuss the generalisation of the
obtained results to other priority models.

%% Below, we will assume that the maximal progress priority $\mu$ is
%% applied after each composition with an interaction model.

@@

\label{property:reset}
Although we define properties as state predicates, any
appropriate logic can be used to specify them.  For instance, the
property \emph{``There is always a possibility to reset the system
  after a single failure''} (\ie without
  additional failures having to occur in the meantime) enforced by the
  Failure Monitor architecture can be specified using CTL as
%  \noteSB{Removed $\lnot \PortResume$, because not needed \emph{and} could not reformulate nicely in English.}
    $\AG
\bigl(\PortFail \rightarrow \EX \EU[]{\lnot \PortFail}{
  \PortReset}\bigr)$.  Additional properties are provided in
\app{properties}.
%% \todoSB{Cite the original paper somewhere.}
An architecture enforces
its characteristic property on  its operand components.
From this point of view, the set of coordinating components is
not relevant, neither are their states.  Thus, 
properties enforced by architectures only involve
the unrestricted composition of the operands:

@@

%\addSB{%
%
\Thm{combining} implies that safe BIP systems can be designed
\emph{compositionally}.  Indeed, it is sufficient to verify that
1)~the applied architectures do enforce their characteristic
properties and 2)~that they are pairwise upwards compatible.  To
a large extent, item~2, \ie verification of the upwards
compatibility, can be carried out syntactically by analysing the
structure of the connectors defining the interaction models:
maximal progress is mostly relevant for connectors involving
triggers; whenever a trigger port is not common to both
architectures, upwards \emph{in}compatibility cannot arise.  The
next section is devoted to the encoding of architectures into
pNets, which addresses item~1 by symbolic verification.
%
%}

@@

\paragraph{Encoding the interaction model}
%% First of all, recall that $\nopri[A, \partition ] \bydef{=}
%% \pNetTuple{(\nopri[C])^{C \in \cC}, \partition, \nopri[\Gamma]}$.
%% Thus, 
The holes in $\nopri[A, \partition]$ are indexed by the elements of
the partition \partition.  For the encoding of our running example, we
take $\partition = \bigl\{\{\PortFail, \PortResume\},\{\PortAsk,
\PortReset\}\bigr\}$.  This corresponds to the intuition that the
dangling ports {\PortFail} and {\PortResume} will be provided by a
%\noteLH{monitored ou monitoring???}
monitored component, whereas {\PortAsk} and {\PortReset} correspond to
the actions provided by the ``environment'' (other components of the
system) that are invoked in case of a persistent failure.  As for
the encoding of the coordinators, in the synchronisation vectors of
\nopri[\Gamma], we will associate Boolean values to the actions
corresponding to these ports.

@@

%\addSB{
We encode all interactions of a connector in one synchronisation vector.  This will allow us to also encode the maximal progress priority model.
%}
We use
the additional Boolean values associated to each port by the encoding
of coordinator components.  For example, observe that the three ports in 
the connector
{\NameTimer.\PortTick
  \trigsynch (\PortFail \trigsynch \NameCtrl.\PortFail)}
form a ``causality chain'': {\NameCtrl.\PortFail} can only
participate in an interaction if the dangling port {\PortFail}
participates, which in turn can only happen if
{\NameTimer.\PortTick} does so.  These dependencies can be rewritten
as Boolean implications $\NameCtrl.\PortFail \Rightarrow \PortFail$
and $\PortFail \Rightarrow \NameTimer.\PortTick$.
%
The conjunction of these two implications can be used as a guard for
the synchronisation vector encoding this connector.

@@

% \noteEM{Dans le paragraph precedent, tu resume le cas des connecteurs sans data, dans les prochains, tu formalise avec data, mais pour le lecteur ``casual'', le changement n'est pas evident... }

Now we lift this encoding to the data-sensitive case. Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then,
%\addSB{
the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
and the encoding of $\Gamma$ is
%}
$\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha \bigl(b_p^{p \in P_x}\bigr)
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%
with $b_p^{p \in P_x}$ fresh Boolean variables, $\alpha$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of all $p$ by corresponding variables $b_p$.

%\addSB{%
%
For the sake of clarity, we simplify the case study encoding in
\fig{FailureTimer:pNet} by reusing the port names of the original
architecture instead of fresh names $\alpha$.  This is made
possible by the fact that each synchronisation vector involves at
most one action of interest (see the properties in
\app{properties}).
%% In the general encoding, the global actions of the
%% synchronisation vectors should be fresh action names.
%% However,
%% for the properties considered in our case study, we only need to
%% exhibit the actions of the holes.  Furthermore, each
%% synchronisation vector involves at most one such action.
%% This
%% allows us to reuse them as the global actions of the
%% synchronisation vectors, simplifying the translation of
%% properties to be verified (\cf \app{properties}).
%
%}\noteLH{ce
  %% paragraphe sonne vraiment note de rapport de recherche, tu
  %% arrives a lui donner un peu d'abstraction/de recul?}\noteSB{Est-ce que c'est mieux comme \c{c}a ? OUI -- Ludo}

@@

\paragraph{Practical experiments.}%\noteSB{J'ai repris ce paragraphe}
%\noteLH{j ai nomme ce paragraphe}
The above encoding provides a mechanism for the symbolic verification
of architectures using our existing tool
\cite{HMZ-FORTE2016QBMZ-AVOCS18} to compute the open automaton
semantics of an open pNet. This tool computes open transitions from
pLTS behaviours and synchronisation vectors of the pNet, then uses an
SMT engine to check satisfiability of their predicates, minimizing the
size of the resulting automaton.
%% The encoding into the SMT language involves the definition of the
%% action and data domains and operators, and the axiomatisation of
%% domain-dependant properties, to allow the SMT engine to be as precise
%% as possible (even if it cannot be complete).
%% The resulting open automaton will later be subject to (symbolic)
%% model-checking and equivalence checking algorithms, themselves
%% requiring automatic proof assistance. \todoEM{trop flou pour le
%%   moment, mais je n'ose pas dire des trucs trop precis... meme si on a
%%   deja une implem de la FH-bisim, elle n'est pas validee!}

In \fig{ArchFailure:OA} we show the full open automaton
obtained from the pNet in \fig{FailureTimer:pNet}.  Due to space limitations, we do not show the
details of the open transitions, but only the assignments of state
variables, some useful parts of the predicates, and the resulting
action; full details can be found in 
\app{full-results}.
%
%\noteSB{Removed ``As an example...'', since that \ex{TimerOT3} was
%  simplified and does not correspond to the real OT.}
%% As an example, $OT_3$ is the one
%% explained in \ex{TimerOT3} (page~\pageref{ex:TimerOT3}).
This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OT 2 and 3 for
{\PortFail}). It looks like the rough behaviour is
reasonable, although a detailed study shows that the property stated
in page~\pageref{property:reset} does not hold, because of the
{\PortFail} loops on states $11$ and $12$. This is because we did not
use the maximal progress assumption here. If we do, we get the
corrected behaviour in \fig{ArchFailure:OA-MaxProgress}
(\app{full-results}), in which $OT_7$ has disappeared, and $OT_2$ and
$OT_{10}$ have been restricted to $b0=\false$. This one verifies all
the properties listed in \app{properties}.

@@

\noindent
%\add{
The following appendices are included for convenience and would be integrated into a research report if the paper is accepted for publication. The first appendix provides previously published complementary information, whereas the subsequent ones give additional formal details complementing architecture definitions, proofs of theorems, and the case study.
%}

@@

\bigskip
\noindent
\verbatimfont{\small}
\begin{verbatim}
SV0: <tick(z, t, z1), fail(zone, b1), fail(b0), _> --> fail(b0),
         [(b1->b0)/\(z1=z |Intersection| ((b1?zone:TopInt)+t))]
SV1: <cancel(b1), resume(b2), resume, _> --> resume, (b1=b2)
SV2: <timeout, timeout, _, ask> --> ask
SV3: <_, reset, _, reset> --> reset
SV4: <init, init, _, _> --> init
pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [(t<GET_MAX z)], { t := (t+1) z := z1}
- Transition: t1---cancel(false)-->t1, {}
- Transition: t1---cancel(true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := (t+1)}
pLTS: Control
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false)-->s1, {}
- Transition: s1---fail(zone, false)-->s1, {}
- Transition: s1---fail(zone, true)-->s2, {}
- Transition: s2---fail(zone, false)-->s2, {}
- Transition: s2---resume(false)-->s2, {}
- Transition: s2---resume(true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false)-->s3, {}
- Transition: s3---resume(false)-->s3, {}
- Transition: s3---reset-->s1, {}
Holes: B
Holes: E
\end{verbatim}
\bigskip

@@

\bigskip
\verbatimfont{\small}
\begin{verbatim}
   {t0---init-->t1, s0---init-->s1},   {},  [true],
   {t := 0, z := TopInt, zone := minZone,maxZone}
OT1 ------------------------------------------------------------
    <t0_s0>-----init----><t1_s1>

@@

INFO  BIPArchFailureTimerMax - Total: 12 SATISFIABLE OTs
INFO  BIPArchFailureTimerMax -        235 UNSATISFIABLE OTs
INFO  BIPArchFailureTimerMax - Total Time: 1293 ms
\end{verbatim}
\bigskip

@@

\bigskip
\noindent
$  OT_1  = \openrule
{\emptyset, \{t:=0, z:=TopInt, zone:=[minZone,maxZone]\} }
{\ostate{00} \OTarrow{init} \ostate{11}}
$

@@

  \paragraph{Properties.}

@@

\bigskip
\small\begin{verbatim}
SV0: <tick(z, t, z1), fail(zone, bf, b1), fail(b0), _> --> fail(b0),
         [(b1=>b0)/\((bf/\b0)=>b1)/\(z1=(z |Intersection| ((b1?zone:TopInt)+t)))]
SV1: <cancel(bc, b1), resume(br, b2), resume, _>-->resume, (b1=b2)/\((bc/\br)=>(b1/\b2))
SV2: <timeout, timeout, _, ask> --> ask
SV3: <_, reset, _, reset> --> reset
SV4: <init, init, _, _> --> init

pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
- Transition: t1---cancel(true, false)-->t1, {}
- Transition: t1---cancel(true, true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}
pLTS: Control
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false, false)-->s1, {}
- Transition: s1---fail(zone, true, false)-->s1, {}
- Transition: s1---fail(zone, true, true)-->s2, {}
- Transition: s2---fail(zone, false, false)-->s2, {}
- Transition: s2---resume(true, false)-->s2, {}
- Transition: s2---resume(true, true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false, false)-->s3, {}
- Transition: s3---resume(false, false)-->s3, {}
- Transition: s3---reset-->s1, {}
\end{verbatim}
\bigskip

@@

\subsubsection{Scaling-up: composing architectures.}
